Amortization of Sub-programs in Simulators

### 2019-09-02 paper/learning/notes.tex
updated

% neurlps classification
% Primary: Applications
% Secodnary cat:Data, Challenges, Implementations, and Software
% or Hijacking Malaria Simulators with Probabilistic Programming

@@

\section{Introduction}
\label{sec:intro}
Simulators arise in a myriad of scientific and industrial settings; epidemiology, physics, engineering, climate modelling and so forth [add citations].
One interesting class of simulators are population-based simulators, used in epidemiology, reinforcement learning
and financial modelling [add citations].
Performing inference in such population-based simulators is challenging, because we have nested-programs, 
which come in the form of rejection sampling blocks. 
In addition to this, each population member can traverse a number of different paths and 
can interact with other members, in certain instances, which can lead to a combinatorial number of paths 
to explore. 
Equally, in many simulated settings we cannot correctly construct the functional form of the density of the desired program,
since we do not have explicit access to the density of the \emph{program}.
Where density refers to the joint density constructed by the program, which in this case is represented by the simulator.
This inability to evaluate the joint density has spawned research into how to perform 
inference in likelihood free settings~\cite{greenberg2019automatic, mohamed2016learning, papamakarios2018sequential,brehmer2018mining}
for conditional density estimation. However, whilst many of these methods look to generate a surragote 
for the entire simulator, we propose amortizing only computational expensive sub-components 
of the simulator, which we can automate with probabilistic programming. 
This enables three things. 
1) It reduces the loss of information, as when simulators are amortized with in current
settings we require, a potentially infinite number of samples, to provide a useful learnt representation of the simulator. 
2) It increases computational efficiency, whilst maintaining expressivity. 
3) It converts expensive rejection sampling loops into to learnt functions that can be 
easily sampled from and used to construct a functional form of the likelihood, that enables the 
target density to be fully constructed, enabling one to leverage a large class of sample efficient inference schemes, namely MCMC.

@@

Sub-programs, all called nested programs~\cite{}
To construct the mathematical framework in this setting we employ the notation of~\cite{rainforth2017automating}, 
where \emph{raw samples} are represented by $ f_{a_i}(x_i | \eta)$ and can be thought of as prior terms.

\textbf{To finish - Linked back to figure 1}
In order to correctly characterize this problem we first provide a useful diagram demonstrating 
the density of the system~\ref{fig:systemoverview}. \textbf{Lots to do}

@@

One common problem in traditional simulators, \emph{programs}, outside of the standard probabilistic
programming framework is that they rely heavily on rejection sampling 
to generate samples from the target density. Not only is rejection sampling 
computational inefficient, [especially if criteria in which samples are generated
has been poorly calibrated] (may leave out this statement as it may make a reviewer
question why would amortization work any better.), but it creates complications for inference.
As the rejection sampling loops provide no explicit form of the target density,
we can not simply evaluate the density and leverage statistically and computationally
efficient algorithms such as MCMC sampling. This inhibits our ability to transform an arbitrary
stochastic simulator into a probabilistic programming framework~\cite{baydin2018efficient}. 
To this end, we derive both a mathematical and engineering based framework to facilitate 
the amortization of rejection sampling loops in arbitrary simulators, which provides a path
way to both leverage a larger, more efficient class of inference methods and convert arbitrary
stochastic simulators into probabilistic programs systems.

@@

However, in many of the non-standard ABC forms of the rejection samplers, the acceptance 
probability is completely intractable, or is very difficult to calculate. 
Thus, as useful rejection sampling is for generating samples from densities that we 
would not be able to sample from, it has many short comings. 
These are of course well known and it is the reason for developing more sophisticated 
sampling schemes, such as importance sampling based methods.
However, in the setting of such complex simulators, written over many years and
in convoluted code bases, there is no easy way to re-implement such sampling 
schemes.
Hence, to leverage existing frameworks we can make use of the probabilistic 
programming executation (PPX) protocols~\cite{baydin2018efficient}. 
These protocols allow one to exact all raw sample calls $f_{a_j}$ from the
simulator. This then enables one to easily construct a prior for the full program
$P(\mathcal{T} = x_{1:n_{x}}| \lambda)$. 
In addition to this, once such rejection schemes, \emph{nested programs}, are located, we have a number of 
primatives that we can leverage to isolate calls to those addresses during training, to collect the samples generated during 
\texttt{.forward()} runs of the simulator, 
and during the inference procedure replace calls to those addresses $a_j$ with a
learnt proposal $\gamma_{a_j}(x_j ; \eta_{a_j}, \phi_j ) $.

@@

\section{Experiments: Amortizing Stochastic Processes}
\label{sec:experiments}
\paragraph{TODO list}
\begin{itemize}
  \item Implement baselines
  \item Implement plotting scripts and evaluation scripts
  \item testing code and infrastructure
  \item litrature review on any methods to learn rejection sampling methods, 
  and conditional density estimation. See this paper for kernel approximation 
  methods~\cite{erraqabi2016pliable}. In addition to the methods listed in the 
  rejection sampling subsection. 
\end{itemize}

@@

For $R1$ we have one input $z^{r1} \in \mathbb{R}^{1 \times 1}$ and one output $y^{r1} \in \mathbb{R}^{1 \times 1}$.
For $R2$ we have two inputs $z^{r2} \in \mathbb{R}^{2 \times 1}$ and one output $y^{r2} \in \mathbb{R}^{1 \times 1}$.
For training purposes we generate $n_{b}$ batches of $n_{t}$ samples for training. 
This means that $z^{r1}_{batch}\mathbb{R}^{n_{b} \times 1}$. 
For the loss function we use a simple mean squared error loss \textbf{We can try different loss 
functions and optimizers} and use the
neural network, $\Gamma(z, y)$, to learn a $\hat{\mu}$ and $\hat{\sigma}^{2}$ to parametize 
a normal proposal. For $R1$, this means $\Gamma_{r1}(z,y) \rightarrow \hat{\mu_{r1}},  \hat{\sigma_{r1}}^{2}$  
and so the proposal $q(z_{1} | z_{2}) = \mathcal{N}(\hat{\mu_{r1}},  \hat{\sigma_{r1}}^{2})$. 
Once learned we can then efficiently sample from the parameterized normal distribution, which
represents the density of the sub-program represented by $R1$. 
This is then replaced in the simulator, during inference time.

@@

We discussed at length how to correctly define the objective, I have now updated the objective with the one derivied in the revised discussion,
along with additional notes on rejection sampling.

@@

The main fundemental difference is that IC cares about the full global program, 
we only care about the local (sub / nested programs).

@@

In the IC paper you are approximating $p(x|y)$ the \emph{the global} program, where 
as you can view this work as learning an approximation for the \emph{local}, sub-programs
$P^{in}(x|\nu)$ where the nuissince parameters in this paper are represented by $\psi$.

In the notation of the inference compilation paper we essentailly learn $q_{a_{t},i_{t}}(x_t| x_{1:t-1}, y)$
where the indicies are part of the set $S_r$.
 The difference here though, is that we only care about the input at that address, your proposals 
 are condiitoning on everything upto that point of the executation trace $\mathcal{T}$
 in the simulator. 
 The training data represented by (11) in the IC paper is not the training data we use in this paper,
 we only care about the $x^{m}_t$ or $z$'s and $x$'s in this paper (notation is not entirely consitant throuhg out this paper yet)
 But these are the parameters that are generated from raw sample calls within side the RS loop $f_{a_i \in S_r}(\cdot))$.

@@

\subsection{Method 1: Learn the maoritzation of the sub-programs}
That is learn $\gamma_{a_{j}}(\cdot ; \eta_{a_{j}}, \cdot)$ for each program in the set $S_r$.

>>>>

% neurlps classification
% Primary: Applications
% Secondary cat:Data, Challenges, Implementations, and Software
% or Hijacking Malaria Simulators with Probabilistic Programming

@@

\section{Introduction}
\label{sec:intro}
Simulators arise in a myriad of scientific and industrial settings; epidemiology, physics, engineering, climate modelling and so forth [add citations].
One interesting class of simulators are population-based simulators, used in epidemiology, reinforcement learning
and financial modelling [add citations].
Performing inference in such population-based simulators is challenging, because we have nested-programs, 
which come in the form of rejection sampling blocks. 
In addition to this, each population member can traverse a number of different paths and 
can interact with other members, in certain instances, which can lead to a combinatorial number of paths 
to explore. 
Equally, in many simulated settings we cannot correctly construct the functional form of the density of the desired program,
since we do not have explicit access to the density of the \emph{program}.
Where density refers to the joint density constructed by the program, which in this case is represented by the simulator.
This inability to evaluate the joint density has spawned research into how to perform 
inference in likelihood free settings~\cite{greenberg2019automatic, mohamed2016learning, papamakarios2018sequential,brehmer2018mining}
for conditional density estimation. However, whilst many of these methods look to generate a surrogate 
for the entire simulator, we propose amortizing only computational expensive sub-components 
of the simulator, which we can automate with probabilistic programming. 
This enables three things. 
1) It reduces the loss of information, as when simulators are amortized with in current
settings we require, a potentially infinite number of samples, to provide a useful learnt representation of the simulator. 
2) It increases computational efficiency, whilst maintaining expressivity. 
3) It converts expensive rejection sampling loops into to learnt functions that can be 
easily sampled from and used to construct a functional form of the likelihood, that enables the 
target density to be fully constructed, enabling one to leverage a large class of sample efficient inference schemes, namely MCMC.

@@

Sub-programs, also called nested programs~\cite{rainforth2018nesting}, are an integral part of stochastic simulators.
Within stochastic simulators they are used to simulate densities that cannot be sampled from in any known, or efficient way.
Such nested-programs may be multi-tiered, in the sense that there exists several nested structures within a given sub-program. For example, given an outer sub-process $S_{out}$, that contains another sub-process $S_{in_{1}}$,
which itself contains another sub-program, $S_{in_{2}}$, that is $S_{in_2} \in S_{in_1} \in S_{out}$. 
Due to these recurrent dependency structures within simulators performing statistical inference becomes increasingly challenging, part of this
difficulty is due to the exploration of the domain of the density $R^{d}$. 
Because of the nested structures the inference strategy see a subset of the domain. 
This happens due to stochastic control flow within the sub-processes, which means only a subset of paths are available 
during the evaluation of the density. 
\textbf{Talk about why this causes trouble for the inference procedures etc}

Due to these complexities and the increase in computation time for both inference, \texttt{.backward()}, and the simulation, \texttt{.forward()},
we leverage the use of amortization schemes to, as we show in Section~\ref{sec:amortizesubp}, correctly utilise MCMC inference procedures and reduce
computation time.

Each nested process representing a different density, each simulating some statistical process.

% To construct the mathematical framework in this setting we employ the notation of~\cite{rainforth2017automating}, 
% where \emph{raw samples} are represented by $ f_{a_i}(x_i | \eta)$ and can be thought of as prior terms.

% \textbf{To finish - Linked back to figure 1}
% In order to correctly characterize this problem we first provide a useful diagram demonstrating 
% the density of the system~\ref{fig:systemoverview}. \textbf{Lots to do}

@@

\label{sec:amortizesubp}
One common problem in traditional simulators, \emph{programs}, outside of the standard probabilistic
programming framework is that they rely heavily on rejection sampling 
to generate samples from the target density. 
Not only is rejection sampling 
computational inefficient, [especially if the criteria in which samples are generated
has been poorly calibrated] (may leave out this statement as it may make a reviewer
question why would amortization work any better.), but it creates complications for inference.
As the rejection sampling loops provide no explicit form of the target density,
we can not simply evaluate the density and leverage statistically and computationally
efficient algorithms such as MCMC sampling. This inhibits our ability to transform an arbitrary
stochastic simulator into a probabilistic programming framework~\cite{baydin2018efficient}. 
To this end, we derive both a mathematical and engineering based framework to facilitate 
the amortization of rejection sampling loops in arbitrary simulators, which provides a path
way to both leverage a larger, more efficient class of inference methods and convert arbitrary
stochastic simulators into probabilistic programs systems.

@@

However, in many of the non-standard ABC forms of the rejection samplers, the acceptance 
probability is completely intractable, or is very difficult to calculate. 
Thus, as useful rejection sampling is for generating samples from densities that we 
would not be able to sample from, it has many short comings. 
These are of course well known and it is the reason for developing more sophisticated 
sampling schemes, such as importance sampling based methods.
However, in the setting of such complex simulators, written over many years and
in convoluted code bases, there is no easy way to re-implement such sampling 
schemes.
Hence, to leverage existing frameworks we can make use of the probabilistic 
programming execution (PPX) protocols~\cite{baydin2018efficient}. 
These protocols allow one to exact all raw sample calls $f_{a_j}$ from the
simulator. This then enables one to easily construct a prior for the full program
$P(\mathcal{T} = x_{1:n_{x}}| \lambda)$. 
In addition to this, once such rejection schemes, \emph{nested programs}, are located, we have a number of 
primitives that we can leverage to isolate calls to those addresses during training, to collect the samples generated during 
\texttt{.forward()} runs of the simulator, 
and during the inference procedure replace calls to those addresses $a_j$ with a
learnt proposal $\gamma_{a_j}(x_j ; \eta_{a_j}, \phi_j ) $.

@@

\section{Experiments: Amortizing Stochastic Processes}
\label{sec:experiments}
\paragraph{TODO list}
\begin{itemize}
  \item Implement baselines
  \item Implement plotting scripts and evaluation scripts
  \item testing code and infrastructure
  \item literature review on any methods to learn rejection sampling methods, 
  and conditional density estimation. See this paper for kernel approximation 
  methods~\cite{erraqabi2016pliable}. In addition to the methods listed in the 
  rejection sampling subsection. 
\end{itemize}

@@

For $R1$ we have one input $z^{r1} \in \mathbb{R}^{1 \times 1}$ and one output $y^{r1} \in \mathbb{R}^{1 \times 1}$.
For $R2$ we have two inputs $z^{r2} \in \mathbb{R}^{2 \times 1}$ and one output $y^{r2} \in \mathbb{R}^{1 \times 1}$.
For training purposes we generate $n_{b}$ batches of $n_{t}$ samples for training. 
This means that $z^{r1}_{batch}\mathbb{R}^{n_{b} \times 1}$. 
For the loss function we use a simple mean squared error loss \textbf{We can try different loss 
functions and optimizers} and use the
neural network, $\Gamma(z, y)$, to learn a $\hat{\mu}$ and $\hat{\sigma}^{2}$ to parameterize 
a normal proposal. For $R1$, this means $\Gamma_{r1}(z,y) \rightarrow \hat{\mu_{r1}},  \hat{\sigma_{r1}}^{2}$  
and so the proposal $q(z_{1} | z_{2}) = \mathcal{N}(\hat{\mu_{r1}},  \hat{\sigma_{r1}}^{2})$. 
Once learned we can then efficiently sample from the parameterized normal distribution, which
represents the density of the sub-program represented by $R1$. 
This is then replaced in the simulator, during inference time.

@@

We discussed at length how to correctly define the objective, I have now updated the objective with the one derived in the revised discussion,
along with additional notes on rejection sampling.

@@

The main fundamental difference is that IC cares about the full global program, 
we only care about the local (sub / nested programs).

@@

In the IC paper you are approximating $p(x|y)$ the \emph{the global} program, where 
as you can view this work as learning an approximation for the \emph{local}, sub-programs
$P^{in}(x|\nu)$ where the nuisance parameters in this paper are represented by $\psi$.

In the notation of the inference compilation paper we essentially learn $q_{a_{t},i_{t}}(x_t| x_{1:t-1}, y)$
where the indices are part of the set $S_r$.
 The difference here though, is that we only care about the input at that address, your proposals 
 are conditioning on everything up to that point of the execution trace $\mathcal{T}$
 in the simulator. 
 The training data represented by (11) in the IC paper is not the training data we use in this paper,
 we only care about the $x^{m}_t$ or $z$'s and $x$'s in this paper (notation is not entirely consistent through out this paper yet)
 But these are the parameters that are generated from raw sample calls within side the RS loop $f_{a_i \in S_r}(\cdot))$.

@@

\subsection{Method 1: Learn the amortization of the sub-programs}
That is learn $\gamma_{a_{j}}(\cdot ; \eta_{a_{j}}, \cdot)$ for each program in the set $S_r$.
